Inferring stellar rotation periods using Gaussian processes

### 2017-01-23 documents/ms.tex

\newcommand{\columbia}{1}
\newcommand{\oxford}{2}
\newcommand{\uw}{3}
\newcommand{\princeton}{4}
\newcommand{\naigrain}{333}
\newcommand{\nmcquillan}{100}
\newcommand{\nkois}{1114}
\newcommand{\nkoimcq}{271}
\newcommand{\nkoigood}{222}
\newcommand{\nkoibad}{49}
\newcommand{\kepexample}{1430163}
\newcommand{\kepexampleperiod}{4}
\newcommand{\aigrainexampleperiod}{20.8}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\acfRMS}{9.27}
\newcommand{\pgramRMS}{1.25}
\newcommand{\mcmcRMS}{0.46}

@@

In order to manage the great flexibilty of the GP model and focus on 
reliably retrieving the period parameter, we also impose priors
on the non-period GP parameters that we 
learned during the development of this method.  
Beginning with very broad priors (uniform in log space between 
-20 and 20) on all the parameters, we first explored what the 
typical values of the non-period parameters tended to be when the 
injected periods were successfully recovered.  We also experimented with
constraining the allowed ranges of the parameters, after discovering 
that some regions of parameter space (such as large values of $A$ and $\Gamma$
and small values of $l$) tended to allow fits that ignored the desired
periodicity.  The priors and bounds we used for this work are listed in
table \ref{tab:priors}.  We note that we do not have any specific
quantitative justification for the use of these priors,
except that they produce good results for this particular test set of 
light curves.  We thus caution that other data sets might have different
noise properties, potentially requiring modifications to these priors
on the non-period parameters.  For period, we tried two different methods:
an uninformed (log-flat) prior between 0.5\,d and 100\,d, and an ACF-informed
prior (details in Appendix.)

% We note that our default prior for period is 
% log-uniform between the bounds, but that we were also able to 
% further improve our results by using an ACF-inspired prior.

To sample the posterior in a way that is sensitive to potential 
multimodality, we use the \texttt{emcee3} MCMC sampler.
We initialize 500 walkers with random samples from the prior 
 and use a mix of different proposal distributions: 
specifically, a 2:2:1 mix of KDE, differential evolution, and DESnooker moves. 
[DFM: more words/citations here?]  
We run fifty steps of the sampler at a time, checking for convergence after 
each iteration, up to a maximum of fifty iterations.  We declare convergence if 
the total effective chain length is at least 8$\times$ the maximum autocorrelation
time.  When convergence is achieved, we drop the first two autocorrelation 
lengths in the chain as a burn-in, and randomly choose 5000 samples as 
representative of the posterior.  This fitting process takes several hours 
for a typical simulated light curve, though in some cases it may take 12
hours or longer to converge.

@@

\begin{figure*}
\begin{center}
% \includegraphics[width=6in, clip=true]{figures/compare_mcmc.pdf}
\includegraphics[width=6in, clip=true]{figures/comparison_emcee3_13_12_2.png}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique.
Since the posterior PDFs of rotation periods are often non-Gaussian,
the points plotted here are the highest likelihood samples.
The uncertainties are the 16th and 84th percentiles.}
\label{fig:compare_mcmc}
\end{center}
\end{figure*}

>>>>

\newcommand{\columbia}{1}
\newcommand{\oxford}{3j}
\newcommand{\uw}{4}
\newcommand{\princeton}{2}
\newcommand{\naigrain}{333}
\newcommand{\nmcquillan}{100}
\newcommand{\nkois}{1114}
\newcommand{\nkoimcq}{271}
\newcommand{\nkoigood}{222}
\newcommand{\nkoibad}{49}
\newcommand{\kepexample}{1430163}
\newcommand{\kepexampleperiod}{4}
\newcommand{\aigrainexampleperiod}{20.8}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\acfRMS}{9.27}
\newcommand{\pgramRMS}{1.25}
\newcommand{\mcmcRMS}{0.46}

@@

In order to manage the great flexibilty of the GP model and focus on 
reliably retrieving the period parameter, we also impose priors
on the non-period GP parameters that we
learned during the development of this method.
Beginning with very broad priors (uniform in log space between
-20 and 20) on all the parameters, we first explored what the
typical values of the non-period parameters tended to be when the
injected periods were successfully recovered.  We also experimented with
constraining the allowed ranges of the parameters, after discovering
that some regions of parameter space (such as large values of $A$ and $\Gamma$
and small values of $l$) tended to allow fits that ignored the desired
periodicity.  The priors and bounds we used for this work are listed in
table \ref{tab:priors}.  We note that we do not have any specific
quantitative justification for the use of these priors,
except that they produce good results for this particular test set of
light curves.  We thus caution that other data sets might have different
noise properties, potentially requiring modifications to these priors
on the non-period parameters.  For period, we tried two different methods:
an uninformed (log-flat) prior between 0.5\,d and 100\,d, and an ACF-informed
prior (details in Appendix.)
% We note that our default prior for period is
% log-uniform between the bounds, but that we were also able to
% further improve our results by using an ACF-inspired prior.

To sample the posterior in a way that is sensitive to potential
multimodality, we use the \texttt{emcee3} MCMC sampler.
We initialize 500 walkers with random samples from the prior
 and use a mix of different proposal distributions:
specifically, a 2:2:1 mix of KDE, differential evolution, and DESnooker moves.
[DFM: more words/citations here?]
We run fifty steps of the sampler at a time, checking for convergence after
each iteration, up to a maximum of fifty iterations.  We declare convergence if
the total effective chain length is at least 8$\times$ the maximum autocorrelation
time.  When convergence is achieved, we drop the first two autocorrelation 
lengths in the chain as a burn-in, and randomly choose 5000 samples as 
representative of the posterior.  This fitting process takes several hours 
for a typical simulated light curve, though in some cases it can take 12
hours or longer to converge.

@@

\begin{figure*}
\begin{center}
% \includegraphics[width=6in, clip=true]{figures/compare_mcmc.pdf}
\includegraphics[width=6in, clip=true]{figures/comparison_emcee3_17_1.png}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique.
Since the posterior PDFs of rotation periods are often non-Gaussian,
the points plotted here are the highest likelihood samples.
The uncertainties are the 16th and 84th percentiles.}
\label{fig:compare_mcmc}
\end{center}
\end{figure*}
